---
title: "01B plot REVAL results"
author: "Veronica Mandelli"
output:
  html_document:
    df_print: paged
---
# VISUALIZE REVAL RESULTS
this script analyzes the results from reval clustering on NDA dataset. 
It help visualize: stratification model reproducibility and clusters'characteristics 
```{r,message=FALSE,warning=FALSE}
# load libraries
library(gridExtra)
library(ggplot2)
library(reshape2)
library(wesanderson)
library(MetBrewer)

#library(here)

# set paths
work_dir = '/Users/vmandelli/OneDrive - Fondazione Istituto Italiano Tecnologia/vineland_proj_edition'#here()

plot_path = file.path(work_dir,'plots')
data_path = file.path(work_dir,'results')
results_path = file.path(work_dir,'results')
code_path = file.path(work_dir,'code','R')
source(file.path(code_path,"cohens_d.R")) 

# data
datafile = 'VABS_withcluster_edition_041022.csv'
data <- read.csv(file.path(data_path,datafile))

#check if there are duplicated in the subjects (some longitudinal??)
length(data$subjectkey)-length(unique(data$subjectkey)) # if =0 ok
```

## rename the clusters based on levels
cluster_domain are provided by reval as numeric lables. We then convert them to a verbal/descriptive label:  


```{r,warning=FALSE,error=FALSE }
for (i in (1:length(data$cluster_domain))) {
  if (data[i, 'cluster_domain'] == '1'){
    data[i,'clust_dom'] = "high"
  }
  else if (data[i, 'cluster_domain'] == '0'){
    data[i,'clust_dom'] = "med"
  }
  else if (data[i, 'cluster_domain'] == '2'){
    data[i,'clust_dom'] = "low"
  }
}

# make the cluster column as a factor
data$clust_dom = as.factor(data$clust_dom)
data$clust_dom <- factor( data$clust_dom,levels = c('high','med','low'));
write.csv(table(data$clust_dom), file.path(results_path,'NDA_cluster_reval_numerosity.csv'))
```

## produce scatter plot for the input features - VABS domain standard scores
```{r,message=FALSE, warning=FALSE}
# define colors
pink = c('#980043',"#DD1C77","#C994C7","#E7E1Ef")
blue = c('#08519c',"#3182bd","#9ecae1","#deebf7")
green = c('#006d2c','#31a354','#a1d99b','#e5f5e0')
yellow = c('#ff9900','#ffcc33','#ffff00','#ffff99')
col_list = list(pink,blue,green,yellow)

#color = c("#FF9933","#3399CC","#009900","#CC0099") # old color

# set some useful lists of variables
dom_list = c('high','med','low')
var_list = c('original_communicationdomain_totalb', 'original_livingskillsdomain_totalb', 'original_socializationdomain_totalb','original_motorskillsdomain_totalb')
title_list = c('communication', 'livingskills', 'socialization','motorskills')

# initialize some other useful variables
title = title 
for (db_type in c('tr','ts')){

data2use= subset(data,data$TR_TS== db_type)
plt <- list()

# looping on VABS variables to 
for (j in (1:4)){
  #define varible, color to use and titels to the plots
  var2use = var_list[j]
  col_name = col_list[j]
  color = col_list[[j]] 
  title = title_list[j]
  
  p = ggplot(data = data2use, 
             aes_string(x = 'clust_dom',
                        y = var2use,
                        color='clust_dom')) 
  p = p + geom_jitter() + geom_boxplot(fill = NA, colour = "#000000", outlier.shape = NA)
  p = p + scale_colour_manual(values = color)
  p = p + guides(colour = FALSE)
  p = p + xlab("") + ylab("")
  p = p + ggtitle(title_list[j]) 
  p = p + theme_bw() 
  p = p + theme(plot.title = element_text(hjust = 0.5))
  ## changed here
  p = p + theme(axis.text.x = element_text(face="bold", size=10))
  p = p + theme(axis.text.y = element_text(face="bold", size=10))
  p = p + scale_y_continuous( limits=c(10, 120))
  p = p + theme(axis.text=element_text(size=10),
                axis.title=element_text(size=8,face="bold"))
  
  print(p)
  # evantually save the plot
  #ggsave(paste(plot_path,'/',db_type,'_total_',var2use,".pdf", sep=""))
  plt[[j]] = p
}

# merge the plots form the same variable
# evantually save the plot
#pdf(paste(plot_path,'/',db_type,'_summary_colorixmike.pdf', sep=""))
g = grid.arrange(
  plt[[1]],
  plt[[2]],
  plt[[3]],
  plt[[4]],
  nrow = 1)
#dev.off()
ggsave(paste(plot_path,'/',db_type,"summary.pdf", sep=""),g,height = 3,width = 6)
 }


```

## plot heatmap of Cohen's d
```{r, warning=FALSE, message=FALSE}
heat = list()

for (db_type in c('tr','ts')){

data2use= subset(data,data$TR_TS== db_type)
for (i in (1:4) ){
  var2use = var_list[i]
  print(var2use)
  color = col_list[[i]]
  title = title_list[i]
  
  vector <- c()
  index <- 0
  group_order <- c("high", "med","low")
  # compute cohens'd
  for (A in group_order){
    for (B in group_order){
      index <- index + 1
      C_d <- cohens_d(data2use[data2use$clust_dom==A,var2use],
                      data2use[data2use$clust_dom==B,var2use], 
                      SIGN=FALSE)
      vector[index] <- round(C_d,digits=2)
    }}
  
  m = matrix(vector,3)
  print(m)
  
  ggheatmap = ggplot(melt(m), aes(Var1,Var2, fill=value)) + 
    scale_x_discrete(name="Vabs_clus",limits=c(1,2,3),labels=c("low", "med","high")) + 
    scale_y_discrete(name="Vabs_clus",limits =c(1,2,3),labels=c("low", "med","high")) + 
    geom_raster() +
    scale_fill_gradient2(low = color[4], high = color[1],  limit = c(0,4), guide = "colourbar",name=NULL) +
    theme_minimal()#+
  theme(axis.text.x = element_blank())#element_text(angle = 45, vjust = 1, 
  #size = 14, hjust = 1)) +
  #coord_fixed()
  
  ggheatmap = ggheatmap +
    geom_text(aes(Var1, Var2, label = value), color = "black", size = 25) +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid.major = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank(),
      axis.ticks = element_blank())+
    guides(fill = guide_colorbar(barwidth = 1, barheight = 10,
                                 title.position = "top", title.hjust = 0, title="d"))
  
  # show here the heatmaps to check them
  #print(ggheatmap)
  # eventually save them on a pdf file
   ggsave(paste(plot_path,'/',db_type,'_heatmap_',var2use,".pdf", sep=""),width = 10,height = 9)
  heat[[i]]=ggheatmap
}}
```

### plot REVAL stability plot
```{r,warning=FALSE,error=FALSE}
# read reval output
reval_val = read.csv(file.path(data_path,"val_metric_for_plot_041022.csv"))
reval_val$cat = 'validation'
reval_tr = read.csv(file.path(data_path,"train_metric_for_plot_041022.csv"))
reval_tr$cat = 'train'

# plot them
mat = rbind(reval_tr,reval_val)
colors = c("#cc0000","#000099")
# spaghetti plot
gg.spaghetti_x<-ggplot(mat, aes (x = k, y =value , group = cat, color = cat))
gg.spaghetti_x <-gg.spaghetti_x +geom_line() + geom_point()   +
  geom_errorbar(aes(ymin= value - CI, ymax= value + CI), width = 0.1, size = 0.5) + 
  theme(axis.text.x = element_text(size=10), axis.text.y = element_text(size=10)) +
  scale_color_manual(values=colors)+theme_classic()


 ggsave(file.path(plot_path,'REVAL_stability.png'),width = 6,height = 4)
gg.spaghetti_x
```

### other information about the database NDA useful to be inserted in the paper
plot total database age-sex info - from now train and test will be merged
```{r, warning=FALSE,error=FALSE, include=FALSE}

data2plot = data
data2plot$sex <- factor( data2plot$sex,levels = c("M","F"))

#plot age sample distribution
p<-ggplot(data2plot, aes(x=interview_age, fill=sex, color=sex)) +
  geom_histogram(position="identity", alpha=0.4,bins = 30)+
  geom_vline(aes(xintercept=mean(interview_age)), color="blue",linetype="dashed")+
  scale_color_manual(values=c("#99ccff","#ffcccc"))+
  scale_fill_manual(values=c("#99ccff","#ffcccc"))+
  scale_x_continuous( limits=c(12,72),breaks=c(12, 24,36,48,60,72))+
  scale_y_continuous( limits=c(0,100),breaks=c(0,20,40,60,80,100))+
  labs(title="NDAR database",x="age (months)", y = "sample size")+
  theme_classic()+coord_fixed(ratio = 1)

p
ggsave(file.path(plot_path,'NDAR_samplesize_100.png'),width = 6,height = 4)

# save some demographics tables
sex_NDAR_table = as.data.frame(table(data2plot$sex))
write.csv(sex_NDAR_table,file.path(results_path,'sex_NDAR_table.csv'))
age_NDAR_table =as.data.frame(as.matrix(summary(data2plot$interview_age)))
write.csv(age_NDAR_table,file.path(results_path,'age_NDAR_table.csv'))
cluster_distribution = table(data$cluster_domain)
write.csv(cluster_distribution,file.path(results_path,'cluster_distribution_NDAR.csv'))

# save NDA subjectkey
sub_ids_nda = data[,c("subjectkey","collection_id")] #as.data.frame(unique(data$subjectkey))
write.csv(sub_ids_nda,file.path(results_path,'sub_ids_collection_id_nda.csv'))

# total sample size 
table(data$clust_dom)
```

plot age and sex distribution divided by clusters
```{r, warning=FALSE,error=FALSE, include=FALSE}
color = c("#FF9933","#3399CC","#009900")
### old color "#FF6666","#99CC66","#99FFFF","#E7E1Ef")

data2plot = data
# set sex as a factor

data2plot$sex <- factor( data2plot$sex,levels = c("F","M"))
# age plot
p = ggplot(data =data2plot ,aes_string(x = 'clust_dom', y ='interview_age', color='clust_dom') )
p = p + geom_jitter() + geom_boxplot(fill = NA, colour = "#000000", outlier.shape = NA)
p = p + scale_colour_manual(values = color)
p = p + guides(colour = FALSE)
p = p + scale_colour_manual(values = color)
p = p + xlab("") + ylab('months')
p = p + ggtitle('age')
p = p + theme_bw()
p = p + theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))
p = p + theme(axis.text.x = element_text(face="bold", size=10))
p = p + theme(axis.text.y = element_text(face="bold", size=10))
p = p + scale_y_continuous( limits=c(0, 72))
#ggsave(paste(plot_path,'/NDAR',db_type,'_ageT1.pdf', sep=""))

# sex plot
sex =ggplot(na.omit(subset(data2plot, select = c(sex,clust_dom))), aes(x= sex,  group=clust_dom)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5,size=2) +
  labs(y = "Percent",fill = "gender") +
  facet_grid(~clust_dom) +
  scale_y_continuous(labels = scales::percent)+ ylim(0,0.87)
#sex = sex + ggtitle(paste("NDAR_",db_type,"VABS-cluster",sep=''))
sex = sex + ggtitle('Gender distribution')
sex = sex  + theme_bw()
sex = sex  + theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))
sex = sex  + theme(axis.text.x = element_text(face="bold", size=10))
sex = sex  + theme(axis.text.y = element_text(face="bold", size=10))
sex
#ggsave(file.path(plot_path,paste('NDAR',db_type,'_sex.pdf', sep="")))

lay =rbind(c(1,1,2,2,2,2)
)

# merge age and sex in a unique figure
#pdf(paste(plot_path,'/',db_type,'_age_sex.pdf', sep=""),width = 5, height = 3.5 )
new_grid = grid.arrange(
  p,sex,
  layout_matrix = lay)
#dev.off()
ggsave(file.path(plot_path,'age_sex_NDA_trts.png'),new_grid)


# check if there is difference in the percentage on F/M in the clustersgender_table
gender_table =  table(data2plot$sex, data2plot$clust_dom)
chisq <- chisq.test(t(gender_table))
print(chisq)
```

### summary table and difference in age
```{r, warning=FALSE,error=FALSE, include=FALSE}

# pairwise
sub_1_poss = c('high','high','med')
sub_2_poss = c('med','low','low')
age_table_sum =  as.data.frame(matrix(nrow=3,ncol=7))
colnames(age_table_sum) = c( 'sub1','sub2','tstat','df','pvalue',"mean_sub1","mean_sub2")
for (d in 1:3){
  x= subset(data,data$clust_dom ==sub_1_poss[d])
  y= subset(data,data$clust_dom ==sub_2_poss[d])
  t_res <- t.test(x[,"interview_age"],y[,"interview_age"])
  print(t_res)
  age_table_sum[d,'sub1']=sub_1_poss[d]
  age_table_sum[d,'sub2']=sub_2_poss[d]
  age_table_sum[d,'tstat']=t_res$statistic
  age_table_sum[d,'df']=t_res$parameter
  age_table_sum[d,'pvalue']=t_res$p.value
  age_table_sum[d,'mean_sub1']=t_res$estimate[1]
  age_table_sum[d,'mean_sub2']=t_res$estimate[2]
}
write.csv(age_table_sum,file.path(results_path,'NDA_age_diff_tab_young.csv'))

summary_age_young = as.data.frame(matrix(nrow=4, ncol = 5))
colnames(summary_age_young) = c("mean","sd","median","max","min")
rownames(summary_age_young) = c('total','high','med','low')
for (sub in c('high','med','low')){
  summary_age_young[sub,'mean'] = mean(data[data$clust_dom==sub,'interview_age'])
  summary_age_young[sub,'sd'] = sd(data[data$clust_dom==sub,'interview_age'])
  summary_age_young[sub,'median'] = median(data[data$clust_dom==sub,'interview_age'])
  summary_age_young[sub,'max'] = max(data[data$clust_dom==sub,'interview_age'])
  summary_age_young[sub,'min'] = min(data[data$clust_dom==sub,'interview_age'])
}

  summary_age_young['total','mean'] = mean(data$interview_age)
  summary_age_young['total','sd'] = sd(data$interview_age)
  summary_age_young['total','median'] = median(data$interview_age)
  summary_age_young['total','max'] = max(data$interview_age)
  summary_age_young['total','min'] = min(data$interview_age)
  
write.csv(summary_age_young,file.path(results_path,'NDA_summary_age_young.csv')) 

```


other important information as the proportion of VABS edtion used and the % of the collection IDs of NDA studies
```{r, warning=FALSE,error=FALSE, include=FALSE}
# plot some needed features/tables
#load edition VABS table 
datafile = 'edition_subid.csv'
edition <- read.csv(file.path(data_path,datafile))
# merge
data_ = merge(data, edition[,c("edition","subjectkey","interview_age","collection_id")], by = c("subjectkey", "interview_age"), all = TRUE,all.x = TRUE,all.y = FALSE)
# remove duplicates
data_new= data_[!duplicated(data_), ]

# table for edition of VABS used
tb_ed = table(data_new$edition)
tb_ed
write.csv(tb_ed,file.path(results_path,'edition_vabs_NDAR.csv'))

dia_tab = table(data_new$phenotype)
dia_tab
write.csv(dia_tab,file.path(results_path,'diagnosis_vabs_NDAR.csv'))
ee = list(unique(data$phenotype))
write.csv(ee[[1]],file.path(results_path,'diagnosis_vabs_NDAR.txt'))
for (i in 1:length(data_new$subjectkey)){
  if (data_new[i,'interview_age']<6) {
    data_new[i,'age_bin'] = "0-6"
  }else if(data_new[i,'interview_age']>=6 & data_new[i,'interview_age']<=12 ) {
    data_new[i,'age_bin'] = "6-12"
  }else if(data_new[i,'interview_age']>=13 & data_new[i,'interview_age']<=18 ) {
    data_new[i,'age_bin'] = "13-18"
  }else if(data_new[i,'interview_age']>=19 & data_new[i,'interview_age']<=24 ) {
    data_new[i,'age_bin'] = "19-24"
  }else if(data_new[i,'interview_age']>=25 & data_new[i,'interview_age']<=30 ) {
    data_new[i,'age_bin'] = "25-30"
  }else if(data_new[i,'interview_age']>=31 & data_new[i,'interview_age']<=36 ) {
    data_new[i,'age_bin'] = "31-36"
  }else if(data_new[i,'interview_age']>=37 & data_new[i,'interview_age']<=42 ) {
    data_new[i,'age_bin'] = "37-42"
  }else if(data_new[i,'interview_age']>=43 & data_new[i,'interview_age']<=48 ) {
    data_new[i,'age_bin'] = "43-48"
  }else if(data_new[i,'interview_age']>=49 & data_new[i,'interview_age']<=54 ) {
    data_new[i,'age_bin'] = "49-54"
  }else if(data_new[i,'interview_age']>=55 & data_new[i,'interview_age']<=60 ) {
    data_new[i,'age_bin'] = "55-60"
  }else if(data_new[i,'interview_age']>=61 & data_new[i,'interview_age']<=66 ) {
    data_new[i,'age_bin'] = "61-66"
  }else if(data_new[i,'interview_age']>=67 & data_new[i,'interview_age']<=72 ) {
    data_new[i,'age_bin'] = "67-72"
  }}

age_tb = table(data_new$age_bin)
write.csv(t(age_tb),file.path(results_path,'age_bind_vabs_NDAR.csv'))
mean(data_new$interview_age)
sd(data_new$interview_age)

# collection ids %
datafile = 'collection_ids.csv'
collection_ids <- read.csv(file.path(data_path,datafile),sep = ';')
collection_ids =  as.data.frame(collection_ids[!duplicated(collection_ids), ])

col_tab = as.data.frame(table(data_new$collection_id.x)/length(data_new$collection_id.x)*100)
colnames(col_tab) = c('collection_id',"percentage")
col_tab$percentage=round(col_tab$percentage,digits = 2)

collection = as.data.frame(merge(col_tab,collection_ids , by = "collection_id", all = TRUE,all.x = TRUE,all.y = FALSE))
#write.csv(collection,file.path(results_path,'collection_vabs_NDAR.csv'))

```
