---
title: "03 subtyping model application on USCD dataset"
author: "Veronica Mandelli"
date: "10/20/2020"
output: html_document
---

# MODEL APPLICATION ON UCSD DATASET - LONGITUDINAL ANALYSIS OF MSEL AND VABS DEVELOPMENTAL TRAJECTORIES
in this script we are going to analyze longitudinal data from UCSD dataset on MSEL and VABS.
subjecta have been already classified into the 3 VABS subtypes using the script '02_UCSD_classification.ipynb'

```{r, message=FALSE, warning=FALSE , error = FALSE, hide=TRUE}
# load packages
library(easypackages)
library(MetBrewer)
library(emmeans)
library(nlme)
library(wesanderson)
libraries('here','ggplot2','gplots','reshape2','gridExtra')

#define paths
work_dir = '/Users/vmandelli/OneDrive - Fondazione Istituto Italiano Tecnologia/vineland_proj_edition'
code_path = file.path(work_dir,'code','R')
uscd_data_path = file.path(work_dir,'results')
result_path = file.path(work_dir,'results')
plot_path = file.path(work_dir,'plots')
source(file.path(code_path,"spaghettiPlot_new.R"))
source(file.path(code_path,"spaghettiPlot.R"))


source(file.path(code_path,"spaghettiPlot2_nodataset.R"))
color = wes_palette("IsleofDogs1")


```

### load data a preliminary checks
```{r,message=FALSE, warning=FALSE ,}
# load data
data <- read.csv(file.path(uscd_data_path,'ucsd_long_allSubj_clusters_061022.csv'))

#check that we load the correct database (1216 T1 ASD subject should be there)
data_asd= data[data$cluster=='1.0' | data$cluster=='2.0' |data$cluster=='0.0',]
length(data_asd[data_asd$time == 1,'cluster'])

# prepare a table with some descriptive statistic of UCSD dataset
descriptive_table =  as.data.frame(matrix(nrow=5, ncol=5))
rownames(descriptive_table)=c("t1","t2","t3","t4","t5")
colnames(descriptive_table)=c('tot','m','f','age_mean','age_std')

for (i in 1:5){
  data2use = data_asd[data_asd$time == i,]
  descriptive_table[i,'tot'] = sum(!is.na(data2use$mullen_ageMo))
  descriptive_table[i,'m'] = sum(!is.na(data2use[data2use$gender=="M",'mullen_ageMo']))
  descriptive_table[i,'f'] = sum(!is.na(data2use[data2use$gender=="F",'mullen_ageMo']))
  descriptive_table[i,'age_mean'] =  mean(data2use$mullen_ageMo,na.rm = TRUE)
  descriptive_table[i,'age_std'] =  sd(data2use$mullen_ageMo,na.rm = TRUE)
}
#write.csv(descriptive_table,file.path(result_path,'UCSD_timepoints_description_ASD.csv'))

##### TD############################################################################################################

data_td= data[data$cluster=='TD' ,]
length(data_td[data_td$time == 1,'cluster'])

# prepare a table with some descriptive statistic of UCSD dataset
descriptive_table =  as.data.frame(matrix(nrow=5, ncol=5))
rownames(descriptive_table)=c("t1","t2","t3","t4","t5")
colnames(descriptive_table)=c('tot','m','f','age_mean','age_std')

for (i in 1:5){
  data2use = data_td[data_td$time == i,]
  descriptive_table[i,'tot'] = sum(!is.na(data2use$mullen_ageMo))
  descriptive_table[i,'m'] = sum(!is.na(data2use[data2use$gender=="M",'mullen_ageMo']))
  descriptive_table[i,'f'] = sum(!is.na(data2use[data2use$gender=="F",'mullen_ageMo']))
  descriptive_table[i,'age_mean'] =  mean(data2use$mullen_ageMo,na.rm = TRUE)
  descriptive_table[i,'age_std'] =  sd(data2use$mullen_ageMo,na.rm = TRUE)
}
#write.csv(descriptive_table,file.path(result_path,'UCSD_timepoints_description_TD.csv'))

```

## plot sample description (age and sex)
```{r,message=FALSE, warning=FALSE ,}
col2use = c("subjectid","gender","vine_agemo")

data2plot = data_asd[data_asd$time == 1,col2use]
data2plot$gender <- factor(data2plot$gender,levels = c("M","F"))

p<-ggplot(data2plot, aes(x=vine_agemo, fill=gender, color=gender)) +
  geom_histogram(position="identity", alpha=0.4,bins = 30)+
  geom_vline(aes(xintercept=mean(vine_agemo)), color="blue",linetype="dashed")+
  scale_color_manual(values=c("#99ccff","#ffcccc"))+
  scale_fill_manual(values=c("#99ccff","#ffcccc"))+
  scale_x_continuous( limits=c(12,72),breaks=c(12, 24,36,48,60,72))+
  scale_y_continuous( limits=c(0,100),breaks=c(0,20,40,60,80,100))+
  labs(title="UCSD database",x="age (months)", y = "sample size")+
  theme_classic()+coord_fixed(ratio = 1)
  
p
#ggsave(file.path(plot_path,'UCSD_T1_ASD_samplesize.png'),width = 6,height = 4)

# save some demographics
sex_USCD_table = as.data.frame(table(data2plot$gender))
#write.csv(sex_USCD_table,file.path(result_path,'UCSD_sex_ASD.csv'))
age_USCD_table =as.data.frame(as.matrix(summary(data2plot$vine_agemo)))
#write.csv(age_USCD_table,file.path(result_path,'USCD_ageT1_ASD.csv'))
```

### prepare data for the longitudinal plot
```{r,message=FALSE,warning=FALSE}
# prepare data
# column with subject_id should be called subjectId
colnames(data)[which(names(data) == "subjectid")] <- "subjectId"

# selct only subjects that have a cluster/dignosis (exclude those ASD with t1>72, or that do not have longitudinal data)
data$cluster[data$cluster==''] <- NA 
data = data[!is.na(data$cluster),]

# rename row 
row.names(data) <- NULL

#age should be round
data$vine_agemo = round(data$vine_agemo)
data$mullen_ageMo = round(data$mullen_ageMo)

# remane in a more intuitive format
  for (i in (1:length(data$subjectId))) {
    if (data[i, 'cluster'] == "1.0"){
      data[i,'cluster'] = "high"
    }
    else if (data[i, 'cluster'] == "0.0"){
      data[i,'cluster'] = "med"
    }
    else if (data[i, 'cluster'] == "2.0"){
      data[i,'cluster'] = "low"
      }
  }

#cluster calumn should be a factor
#define cluster columns as a factor
data$cluster <- as.factor(data$cluster)

# correct some "0"
data[data == 0] <- NA   

cl_number_tab = table(data$cluster)/5
write.csv(cl_number_tab,file.path(result_path,'UCSD_cluster_samplesize.csv'))
```

### longitudinal analysis
selct here if you want to run it on MSEL or VABS 
```{r}
# variable selection

# MELS
# var2use = c('mullen_VR_AgeEq','mullen_FM_AgeEq','mullen_RL_AgeEq','mullen_EL_AgeEq')
# X_var_ = 'mullen_ageMo'
# title_list= c("MSEL_VR","MSEL_FM","MSEL_RL","MSEL_EL")
# y_lim_ = c(0, 90)
# x_lim_ = c(0, 60)
# post_hoc_folder  = 'MSEL_UCSD_REVAL'
# lme_folder = 'UCSD_MSEL_LONG_REVAL'
# name_lme = 'ANOVA_lme_spagh2_UCSD_REVAL_MSEL_'
# name_posthoc = 'Post_hoc_ANOVA_lme_spagh2_UCSD_MSEL_'
# name_sum_post_hoc_ = "Posthoc_UCSD_MSELlong"
# plotname = "MSEL"
# name4pdf = 'MELS_long_UCSD_reval.pdf'
# 
# #VABS
var2use = c('vine_ComTotal_DomStd','vine_DlyTotal_DomStd','vine_SocTotal_DomStd','vine_MtrTotal_DomStd')
X_var_ = 'vine_agemo'
title_list= c("VABS_comm","VABS_dailyski","VABS_soc","VABS_mot")
y_lim_ = c(25, 125)
x_lim_ = c(0,85)
post_hoc_folder  = 'VABS_UCSD_REVAL'
lme_folder = 'UCSD_VABS_LONG_REVAL'
name_lme = 'ANOVA_lme_spagh2_UCSD_REVAL_VABS_'
name_posthoc = 'Post_hoc_ANOVA_lme_spagh2_UCSD_VABS_'
name_sum_post_hoc_ = "Posthoc_UCSD_VABSlong"
plotname = "VABS"
name4pdf = 'VABS_long_UCSD_reval.pdf'
```

### select data for plotting (including TD) and for statistics (exluding TD)
```{r,message=FALSE,warning=FALSE}
# for plotting

data2plot = data[data$cluster=="high" | data$cluster=="med" |data$cluster=="low"|data$cluster=="TD",]
data2plot$cluster <- factor( data2plot$cluster,levels = c("high","med","low","TD"));

# for statistic
data2stat = data[data$cluster=="high" | data$cluster=="med" |data$cluster=="low",]
data2stat$cluster <- factor( data2stat$cluster,levels = c("high","med","low"));
data2stat$dataset = "UCSD"
```

### STATISTIC: linear mixed effect model only on ASD groups.
```{r, warning=FALSE, error= FALSE}
# size for multiple comparison
size = 3 # 3 HP for each subscale
library(openxlsx)
#using spaghettiplot2
plt = list()
for (i in 1:length(var2use)) {
  col2use =var2use[i]
  title= title_list[i]
  resp = spaghettiPlot2_nodataset(
    data2use = data2stat,
    x_var = X_var_,
    y_var = col2use,
    subgrp_var = "cluster",
    cov_vars=NULL,
    xLabel = 'age',
    yLabel = "age equivalent",
    modelType = "linear",
    fname2save = NULL,
    plot_dots = FALSE, plot_lines = TRUE,
    dot_alpha = 1/10, line_alpha = 1/10,
    xLimits = x_lim_, yLimits = y_lim_,
    lineColor = "cluster", dotColor = "cluster",
    legend_name = "Slope", maxIter = 500,
    plot_title = title
  )
  
  
  #change color 
  cols2use = color
  font_size = 11
  p_com = resp$p
  p_com = p_com + scale_colour_manual(values = cols2use) + scale_fill_manual(values = cols2use)
  p_com = p_com + theme_bw(base_size = font_size) +
    theme(plot.title = element_text(hjust = 0.5, size=font_size)) +
    guides(colour = FALSE, fill = FALSE)
  
  plt[[i]] = p_com
 
  
  # change format to be readable in excel
  lme2save = anova(resp$lme_model)
  
  
  #correct p value for multiple comparison
  for (a in 1:3){
    p = lme2save[a,"Pr(>F)"]
    print('p')
    print(p)
    p_adj_ = p.adjust(p, method = "fdr", n = size)
    print('p_adj_')
    print(p_adj_)
    lme2save[a,'p_adj_temp'] = p_adj_
  }
  
  
  lme_definitive_2save  = lme2save[,c("Sum Sq","Mean Sq","NumDF","DenDF","F value","Pr(>F)","p_adj_temp")]
  colnames(lme_definitive_2save) = c("Sum Sq","Mean Sq","NumDF","DenDF","F value","p-value","p_adj")
  # save
  write.xlsx(lme_definitive_2save,file.path(result_path,lme_folder, paste("ANOVA",name_lme,col2use,".xlsx",sep='')))
}
```

### produce the plots  
```{r, message=FALSE, warning=FALSE ,error= FALSE}
plt_old=list()
#title_list = c("VR","FM","RL","EL")
title_list=c("c","d","s","m")
data2stat$dataset  = "UCSD"
for (i in 1:length(var2use)) {
  
  col2use =var2use[i]
  #plot line
  title = title_list[i]
  # plot 
  spagh =   spaghettiPlot2_nodataset( 
    data2use = data2stat,
    x_var = X_var_,
    y_var = col2use,
    subgrp_var = "cluster",
    cov_vars=NULL,
    xLabel = 'age',
    yLabel = "standard score",
    modelType = "linear",
    fname2save = NULL,
    plot_dots = FALSE, plot_lines = TRUE,
    dot_alpha = 1/10, line_alpha = 1/10,
    xLimits = x_lim_, yLimits = y_lim_,
    lineColor = "cluster", dotColor = "cluster",
    legend_name = "Slope", maxIter = 500,
    plot_title = title
  )
  
  
  # change plot color 
  # change color 
  cols2use = color
  font_size = 11
  p_com = spagh$p
  p_com = p_com + scale_colour_manual(values = cols2use) + scale_fill_manual(values = cols2use)
  p_com = p_com + theme_bw(base_size = font_size) +
    theme(plot.title = element_text(hjust = 0.5, size=font_size)) +
    guides(colour = FALSE, fill = FALSE)
  
  # save plot in a list
  plt_old[[i]] = p_com
  print(p_com)

  # eventually save
  ggsave(file.path(plot_path,paste("UCSD_",plotname,'_',col2use,'.png',sep="")), p_com, width = 4.5 ,height = 4)
  
}


# merge plots

g = grid.arrange(
  plt_old[[1]],
  plt_old[[2]],
  plt_old[[3]],
  plt_old[[4]],
  nrow = 2)
g
ggsave(file.path(plot_path,paste('UCSD_REVAL_',plotname, '.png',sep="")), g, width = 10 ,height = 6)
library(patchwork)
pfd2save = plt_old[[1]] + plt_old[[2]] + plt_old[[3]] + plt_old[[4]]
ggsave(file.path(plot_path,name4pdf), pfd2save, width = 6 ,height = 4)
```

### post hoc comparison for the lme model
```{r, message=FALSE, warning=FALSE ,error= FALSE}
library(emmeans)
# for statistic
data2stat = data[data$cluster=="high" | data$cluster=="med" |data$cluster=="low",]
data2stat$cluster <- factor( data2stat$cluster,levels = c("high","med","low"));
sub_list = c("high","med","low")

# initialize some empty saving matrices
ANOVA_post_hoc = as.data.frame(matrix(ncol=24,nrow=1))

colnames_2save = c("col2use","sub_A","sub_B","SumSq_age","MeanSq_age","NumDF_age","DenDF_age","Fvalue_age","p-value_age",
                   "p_adj_age", "SumSq_cluster","MeanSq_cluster","NumDF_cluster","DenDF_cluster",
                   "Fvalue_cluster","p-value_cluster","p_adj_cluster",
                   "SumSq_age:cluster","MeanSq_age:cluster","NumDF_age:cluster","DenDF_age:cluster",
                   "Fvalue_age:cluster","p-value_age:cluster","p_adj_age:cluster")

colnames(ANOVA_post_hoc) = colnames_2save

# multiple comparison size
size = 3 #number of comparisons for number of VABS scales
sub_poss1 = c("high","high","med")
sub_poss2 = c("med","low","low")

for (i in (1:4)) {
  col2use = var2use[i]
  print(col2use)
  
  # post hoc done by hand!
  for(o in 1:3){
    
    sub1 = sub_poss1[o]
    substypeA =  subset(data2stat,data2stat$cluster==sub1)
    #for (d in 1:3){
    
    sub2 = sub_poss2[o]
    substypeB=  subset(data2stat,data2stat$cluster==sub2)
        # define vars
    
        title= title_list[i]
        x_var = X_var_
        y_var = col2use
        subgrp_var = "cluster"
        maxIter = 500
        df2use = rbind(substypeA,substypeB)
       
        # define the formula
        form2use = as.formula(sprintf("%s ~ %s*%s + (%s|subjectId)", #(1|%s) + 
                                  y_var, x_var, subgrp_var, x_var))# "dataset",
        
        ctrl = lmerControl(optCtrl=list(maxfun=maxIter), # lmerControl(optCtrl=list(maxfun=maxIter)
                     check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4),check.nobs.vs.nRE = "ignore")
        # run the model
        m_allgrps <- eval(substitute(lmer(formula = form2use, data = df2use, na.action = na.omit, control = ctrl)))
        
        # saving 
        lme2save = anova(m_allgrps)
         
          
          #correct p value for multiple comparison
          for (a in 1:3){
            p = lme2save[a,"Pr(>F)"]
            p_adj_ = p.adjust(p, method = "fdr", n = size)
            lme2save[a,'p_adj_temp'] = p_adj_
          }
          
        
        #save the post-hoc lme complete
        lm = lme2save[,c("Sum Sq","Mean Sq","NumDF","DenDF","F value","Pr(>F)","p_adj_temp")]
        colnames(lm) = c("Sum Sq","Mean Sq","NumDF","DenDF","F value","p-value","p_adj")
        write.xlsx(lm,file.path(result_path,"post_hoc",post_hoc_folder,paste(name_posthoc,col2use,"_",sub1,"_",sub2,".xlsx",sep='')))
        
        #save in a unique table
        vect2save = c(col2use,sub1,sub2,as.list(as.vector(t(lm))))
        ANOVA_post_hoc = rbind(ANOVA_post_hoc,vect2save) 
      }
}

write.xlsx(ANOVA_post_hoc[2:length(ANOVA_post_hoc$col2use),],file.path(result_path,paste(name_sum_post_hoc_,".xlsx",sep='')))
```

### SINGLE GROUP PLOTS uncomment here if you want plot single group trajectories only
```{r, message=FALSE, warning=FALSE ,error= FALSE}
# x_axis = 'mullen_ageMo'
# color_list = c("#FF9933","#3399CC","#009900","#CC0099")
# group_list = c("high",'med','low')
# plt_sub = list()
# lay =rbind(c(1,2),c(3,4))
# 
# for (subtype in 1: length(group_list)){
# # select the subtype
# data2use = data[data$cluster== group_list[subtype],]
# data2use$cluster <- factor( data2use$cluster,levels = group_list[subtype]);
#       #select the variables to plot
#       for (var in 1:length(var2use)){
#       col2use = var2use[var]
#       Single = ggplot(data = data2use, aes(x =get(x_axis), y = get(col2use), group = subjectId))
#       Single = Single + geom_line(aes(colour = group_list[subtype] ), alpha = 0.5) + guides(alpha = FALSE)
#       Single = Single + scale_colour_manual(values=color_list[subtype])
#       Single = Single + theme(legend.title=element_blank()) + xlab('vine_agemo') + ylab(col2use)+ xlim(0,60)
#       #print(Single)
#       plt_sub[[var]]=Single
#       }
#      new_grid_sub= grid.arrange(
#      plt_sub[[1]],
#      plt_sub[[2]],
#      plt_sub[[3]],
#      plt_sub[[4]],
#      layout_matrix = lay)
#      new_grid_sub

# }


```

### DOTS AND LINES ONLY PLOTS  uncomment if you want plots only with lines and/or dots ######################################
```{r, message=FALSE, warning=FALSE ,error= FALSE}
######### DOTS AND LINES ONLY PLOTS  uncomment if you want plots only with lines and/or dots ######################################
# plt_p1=list()
# plt_p =  list()
# for (i in 1:length(var2use)) {
#         col2use =var2use[i]
# 
#     spagh = spaghettiPlot_new(df=data2use,
#                           x_var=X_var_
#                           y_var=col2use,
#                           subgrp_var='cluster',
#                           xLabel = 'age',
#                           yLabel =col2use,
#                           modelType = "linear",
#                           fname2save = NULL, plot_dots = FALSE, plot_lines =TRUE,
#                           ci_band = TRUE, pi_band = FALSE, dot_alpha = 1/10, line_alpha = 1/10,
#                           band_alpha = 1/2, xLimits = c(0, 60), yLimits = c(0,90), standardize = FALSE,
#                           #color = c("#999999","#999999","#999999","#999999")
#                           color = c("#333333","#333333","#333333","#333333"))#wes_palette("IsleofDogs1")
#                             #c("#FF9933","#3399CC","#009900","#CC0099")) # ask mike standardized
#     
#     print(spagh$p)
#     print(spagh$p1)
#     plt_p1[[i]] = spagh$p1
#     #ggsave(file.path(plot_path,paste('onlylines_',col2use,'_LINES_col.png',sep="")), spagh$p, width = 6 ,height = 4)
#     plt_p[[i]] = spagh$p
#     print(col2use)
#     #ggsave(file.path(plot_path,paste('FORMIKE_',col2use,'_LINES_summary.png',sep="")), spagh$p, width = 6 ,height = 4)
# }

```


###   get betas from LME (rate of growth)
this part is meaningful only if considering the MSEL.
```{r,message=FALSE, warning=FALSE ,error= FALSE}
# define my subtype list
subtype_list = c("high","med","low")
#initialize an empty matrix to save betas in a comprhensible way
betas_matrix= as.data.frame(matrix(nrow=4,ncol=3))
rownames(betas_matrix)= var2use
colnames(betas_matrix)= subtype_list


# run the nestedfor-loop 
for (k in 1:length(subtype_list)){
   #extracting the betas:
  subtype2use = subtype_list[k]
  subset_group = data2use[data2use$cluster== subtype2use ,]
  for (i in 1:length(var2use)){
        col2use =var2use[i]
        x_var='mullen_ageMo'
        y_var=col2use 
        #subgrp_var = 'hybrid_cluster'
        fx_form = as.formula(sprintf("%s ~ %s", y_var, x_var))   # fixed effect --> delate the subtype because I                                                                                         want betas only from one and not a comparison between them?
        rx_form = as.formula(sprintf("~ %s|subjectId",x_var))                 # random effect
        ctrl <- lmeControl(opt='optim', msMaxIter = 500)
        ml_method = "ML"
        m_allgrps <- eval(substitute(lme(fixed = fx_form, random = rx_form, data = subset_group,
                                     na.action = na.omit, control=ctrl, method = ml_method), 
                                     list(fx_form = fx_form, rx_form = rx_form)))
    
        print (subtype2use)
        print (col2use)
        print(coef(summary(m_allgrps)))
        coef = coef(summary(m_allgrps))[2,1]
        betas_matrix[i,k] = coef
  }
  
}
print(betas_matrix)
#write.csv(betas_matrix,file.path(result_path,'betas_matrix_UCSD_REVAL_MSEL.csv'))
```

### uncomment here if you want to check clusters numerosity and sex distribution
```{r,warning=FALSE,error=FALSE,message=FALSE, include=TRUE}
# take only T1
new_data = data[data$time==1, ]

# take only ASD
asd_only = new_data[new_data$cluster=="high"|new_data$cluster=="med"|new_data$cluster=="low",]
asd_only$cluster <- factor( asd_only$cluster,levels = c('high','med','low'))
asd_only$gender <- factor( asd_only$gender,levels = c('F','M'))


# exclude subjects with incert gender
asd_only = asd_only[asd_only$gender %in% c("M","F"),]
gender_table = table(asd_only$cluster,asd_only$gender)

# check if there is difference in the percentage on F/M in the clusters
# all tre subtypes together
chisq <- chisq.test(t(gender_table))
print(chisq)
print(gender_table)


# there is no difference in the proportion of M/F across subtypes, p>0.05

```

### uncomment here if you want some plots
```{r, include = FALSE,message=FALSE, warning=FALSE ,error= FALSE}
# are there any differencesc in AGE at T1
color = c("#FF9933","#3399CC","#009900","#CC0099")
                    
data2plot = new_data[new_data$cluster=="high" | new_data$cluster=="med" |new_data$cluster=="low",]
data2plot$cluster <- factor( data2plot$cluster,levels = c('high','med','low'));
#age plot
p = ggplot(data =data2plot ,aes_string(x = 'cluster', y = 'vine_agemo', color='cluster') )
p = p + geom_jitter() + geom_boxplot(fill = NA, colour = "#000000", outlier.shape = NA)
p = p + scale_colour_manual(values = color)
p = p + guides(colour = FALSE)
p = p + scale_colour_manual(values = color)
p = p + xlab("") + ylab('months')
p = p + ggtitle('age at T1')
p = p + theme_bw()
p = p + theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))
p = p + theme(axis.text.x = element_text(face="bold", size=10))
p = p + theme(axis.text.y = element_text(face="bold", size=10))
p = p + scale_y_continuous( limits=c(0, 72))
                    
# sex plot
sex =ggplot(na.omit(subset(asd_only, select = c(gender,cluster))), aes(x= gender,  group=cluster)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..),
                   y= ..prop.. ), stat= "count", vjust = -.5,size=2) +
    labs(y = "Percent",fill = "") +
    facet_grid(~cluster) +
    scale_y_continuous(labels = scales::percent)+ ylim(0,0.87)

sex = sex + ggtitle('Gender distribution')
sex = sex  + theme_bw()
sex = sex  + theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))
sex = sex  + theme(axis.text.x = element_text(face="bold", size=10))
sex = sex  + theme(axis.text.y = element_text(face="bold", size=10))
sex

lay =rbind(c(1,1,2,2,2,2)
             )

pdf(file.path(plot_path,'UCSD_age_sex.pdf'),width = 5, height = 3.5 )
new_grid = grid.arrange(
     p,sex,
      layout_matrix = lay)
dev.off()
```

### uncomment here if you want to chek lme results inserting GENDER as covariate
```{r, echo = FALSE, include = FALSE}
# do lme with gender as covariate
# prepare the dataset
data2stat_gender =  data2stat[data2stat$gender %in% c ("M","F"),]
data2stat_gender$gender =  factor(data2stat_gender$gender,levels = c('F','M'))
size = 4
# run lme
for (i in 1:length(var2use)) {
  col2use =var2use[i]
  title= title_list[i]
  resp = spaghettiPlot2_nodataset(
    data2use = data2stat_gender,
    x_var = X_var_,
    y_var = col2use,
    subgrp_var = "cluster",
    cov_vars='gender',
    xLabel = 'age',
    yLabel = "age equivalent",
    modelType = "linear",
    fname2save = NULL,
    plot_dots = FALSE, plot_lines = TRUE,
    dot_alpha = 1/10, line_alpha = 3/10,
    xLimits = x_lim_, yLimits = y_lim_,
    lineColor = "cluster", dotColor = "cluster",
    legend_name = "Slope", maxIter = 500,
    plot_title = title
  )
  
  # change format to be readable in excel
  lme2save = anova(resp$lme_model)
  
  
  #correct p value for multiple comparison
  for (a in 1:4){
    p = lme2save[a,"Pr(>F)"]
    p_adj_ = p.adjust(p, method = "fdr", n = size)
    lme2save[a,'p_adj_temp'] = p_adj_
  }
 
  
  lme_definitive_2save  = lme2save[,c("Sum Sq","Mean Sq","NumDF","DenDF","F value","Pr(>F)","p_adj_temp")]
  colnames(lme_definitive_2save) = c("Sum Sq","Mean Sq","NumDF","DenDF","F value","p-value","p_adj")
  # save
 
  write.xlsx(lme_definitive_2save,file.path(result_path,'GND_ascov',paste(name_lme,col2use,".xlsx",sep='')))
}
```

