---
title: "04 models comparisons REVAL vs VABSnorms"
author: "Veronica Mandelli"
output: html_document
---
# VABS NORMS STRATIFICATION 
in this script we create subtype of autistic adaptive behavior by looking at the VABS total score and compare it with the norms from the VABS manual. 
<br>
* TOTAL VABS>=85 --> the child performs as a neurtypical aged matched peer (high)
* TOTAL VABS< 85 and >=70 --> the child perfroms one standard deviation below the norm (med)
* TOTAL VABS<70 --> the child performs 2 standard deviation below the norm (low)

### load packages, functions, data and define paths
```{r,echo=FALSE, message=FALSE,warning=FALSE}
#load packages 
library(easypackages)
libraries('here','ggplot2','gplots','reshape2','gridExtra','ggalluvial','nlme','emmeans')

# define paths
work_dir = '/Users/vmandelli/OneDrive - Fondazione Istituto Italiano Tecnologia/vineland_proj_new'
code_path = file.path(work_dir,'code','R')
#uscd_data_path = file.path(work_dir,'results','prova4soggetti')
uscd_data_path = file.path(work_dir,'results','FINAL_correct')
ndar_data_path = file.path(work_dir,'results','FINAL_correct')
result_path = file.path(work_dir,'results','FINAL_correct')
plot_path = file.path(work_dir,'plots','FINAL_correct')

#load some fucntions
source(file.path(code_path,"spaghettiPlot_new.R"))
source(file.path(code_path,"spaghettiPlot.R"))
source(file.path(code_path,"spaghettiPlot2_nodataset.R"))
source(file.path(code_path,"alluvial.R"))

#load data
data <- read.csv(file.path(uscd_data_path,'ucsd_long_allSubj_clusters.csv'))
```

### reformat data 
```{r,message=FALSE,warning=FALSE, error=FALSE}
# column with subject_id should be called subjectId
colnames(data)[which(names(data) == "subjectid")] <- "subjectId"

# selct only subjects that have a cluster/dignosis (exclude those ASD with t1>72, or that do not have longitudinal data)
data$cluster[data$cluster==''] <- NA 
data = data[!is.na(data$cluster),]

# rename row 
row.names(data) <- NULL

# age should be round
data$vine_agemo = round(data$vine_agemo)
data$mullen_ageMo = round(data$mullen_ageMo)

# cluster column should be a factor
# remane in a more comprehansible format
# for domain
for (i in (1:length(data$cluster))) {
  if (data[i, 'cluster'] == '2.0'){
    data[i,'cluster'] = "high"
  }
  else if (data[i, 'cluster'] == '3.0'){
    data[i,'cluster'] = "med"
  }
  else if (data[i, 'cluster'] == '1.0'){
    data[i,'cluster'] = "low"
  }
}
data$cluster <- as.factor(data$cluster)

# correct some "0"
data[data == 0] <- NA   

# define variables to use
var2use = c('mullen_VR_AgeEq','mullen_FM_AgeEq','mullen_RL_AgeEq','mullen_EL_AgeEq')
```

### compute clusters from Vineland TD norms
```{r,message=FALSE,warning=FALSE, error= FALSE}
# prepare data only from ASD
data4normativemodel = data[data$cluster=="high" | data$cluster=="med" |data$cluster=="low",]
data4normativemodel$cluster <- factor( data4normativemodel$cluster,levels = c("high","med","low"));

# mask=rowSums(is.na(data4normativemodel[,var2use]))==0
# data4normativemodel = as.data.frame(data4normativemodel[mask,])

subjetc_list = unique(data4normativemodel$subjectId)

# define norma mean and standard deviation
VABSnorms_STD_vine =  15
VABSnorms_MEAN_vine = 100


# assing subject to their norm cluster
# loop on subects
for (sub in (1:length(subjetc_list))){
  subset_ = subset (data4normativemodel,data4normativemodel$subjectId==subjetc_list[sub])
  if (subset_[1,'vine_AdapBehav_DomStd']>=VABSnorms_MEAN_vine - VABSnorms_STD_vine){
    data4normativemodel[data4normativemodel$subjectId==subjetc_list[sub],'VABSnorms_cluster']="high"
  }
  else if (subset_[1,'vine_AdapBehav_DomStd'] <VABSnorms_MEAN_vine - VABSnorms_STD_vine & subset_[1,'vine_AdapBehav_DomStd'] > VABSnorms_MEAN_vine-2*VABSnorms_STD_vine){
    data4normativemodel[data4normativemodel$subjectId==subjetc_list[sub],'VABSnorms_cluster']="med"
  }
  else if(subset_[1,'vine_AdapBehav_DomStd'] <= VABSnorms_MEAN_vine - 2*VABSnorms_STD_vine){
    data4normativemodel[data4normativemodel$subjectId==subjetc_list[sub],'VABSnorms_cluster']="low"
  }
}

data4normativemodel$VABSnorms_cluster <- factor( data4normativemodel$VABSnorms_cluster,levels = c("high","med","low"));

# alluvial plot --> eventually make an alluvial plot to see how subjects change from reval stratification model to VABS norm startification model
data4normativemodel_T1 = data4normativemodel[!duplicated(data4normativemodel$subjectId), ]
VABSnorms_alluvial <- plotalluvial (data4normativemodel_T1,'subjectId','gender','cluster','VABSnorms_cluster', 'REVAL vs VABSnorms')
# print the table
print(table(data4normativemodel_T1$cluster, data4normativemodel_T1$VABSnorms_cluster)) #row REVAL col VABSnorms
```

### lme for VABS norm stratification
Does VABS norms classification produce developmentally meaningful subtypes? 
<br> run linear mixed effect model on MSEL subscales using as stratification variable VABS norms subtype 
```{r,message=FALSE,warning=FALSE, error= FALSE}
# for statistic
data2stat = data4normativemodel[data4normativemodel$VABSnorms_cluster=="high" | data4normativemodel$VABSnorms_cluster=="med" |data4normativemodel$VABSnorms_cluster=="low",]
data2stat$VABSnorms_cluster <- factor( data2stat$VABSnorms_cluster,levels = c("high","med","low"));
sub_list = c("high","med","low")
title_list= c("MSEL_VR","MSEL_FM","MSEL_RL","MSEL_EL")
size = 3


#lme for the 3 groups and save the results
for (i in (1:4)) {
  col2use =var2use[i]
  #print(col2use)
  title= title_list[i]
  resp = spaghettiPlot2_nodataset(
    data2use = data2stat,
    x_var = 'mullen_ageMo',
    y_var = col2use,
    subgrp_var = "VABSnorms_cluster",
    cov_vars=NULL,
    xLabel = 'age',
    yLabel = "age equivalent",
    modelType = "linear",
    fname2save = NULL,
    plot_dots = FALSE, plot_lines = TRUE,
    dot_alpha = 1/10, line_alpha = 3/10,
    xLimits = c(0, 60), yLimits = c(0, 90),
    lineColor = "cluster", dotColor = "cluster",
    legend_name = "Slope", maxIter = 500,
    plot_title = title
  )
  
  # saving
  lme2save = anova(resp$lme_model)
  lme2save$`F value` = round(lme2save$`F value`, digits = 2)
  lme2save$`DenDF` = round(lme2save$`DenDF`, digits = 2)
  lme2save$`Sum Sq`= round(lme2save$`Sum Sq`, digits = 2)
  lme2save$`Mean Sq` = round(lme2save$`Mean Sq`, digits = 2)
  
  for (j in 1:3){
    if (lme2save[j,"Pr(>F)"]<= 0.001){
      lme2save[j,"p-value_new"] = '<0.001'
    }else if (lme2save[j,"Pr(>F)"]> 0.001){
      #lme2save[j,"p-value_new"] = as.character(as.numeric(round(lme2save[j,"Pr(>F)"],digits = 3)),length= 5)}
      lme2save[j,"p-value_new"] = paste("=",as.character(round(lme2save[j,"Pr(>F)"],digits = 3),sep = ""))}
    
  }
  
  #correct p value for multiple comparison
  for (a in 1:3){
    p = lme2save[a,"Pr(>F)"]
    #print('p')
    #print(p)
    p_adj_ = p.adjust(p, method = "fdr", n = size)
    #print('p_adj_')
    #print(p_adj_)
    lme2save[a,'p_adj_temp'] = p_adj_
  }
  #change p-adj format to be readable in excel
  #print(lme2save)
  for (d in 1:3){
    if (lme2save[d,"p_adj_temp"]<= 0.001){
      lme2save[d,"p_adj"] = '<0.001'
    }else if (lme2save[d,"p_adj_temp"]> 0.001){
      lme2save[d,"p_adj"] = paste("=",as.character(round(lme2save[d,"p_adj_temp"],digits = 3),sep = ""))}
    
  }
  
  lme_definitive_2save  = lme2save[,c("Sum Sq","Mean Sq","NumDF","DenDF","F value","p-value_new","p_adj")]
  colnames(lme_definitive_2save) = c("Sum Sq","Mean Sq","NumDF","DenDF","F value","p-value","p_adj")
  
  
  write.csv(lme_definitive_2save,file.path(result_path,paste('ANOVA_lme_spagh2_UCSD_VABSnorm_MELS_',col2use,".csv")))
  
}
```  

### post hoc comparison for the lme model
```{r,message=FALSE, error = FALSE, warning=FALSE}
# initialize some empty saving matrices
ANOVA_post_hoc = as.data.frame(matrix(ncol=24,nrow=1))

colnames_2save = c("col2use","sub_A","sub_B","SumSq_age","MeanSq_age","NumDF_age","DenDF_age","Fvalue_age","p-value_age",
                   "p_adj_age", "SumSq_cluster","MeanSq_cluster","NumDF_cluster","DenDF_cluster",
                   "Fvalue_cluster","p-value_cluster","p_adj_cluster",
                   "SumSq_age:cluster","MeanSq_age:cluster","NumDF_age:cluster","DenDF_age:cluster",
                   "Fvalue_age:cluster","p-value_age:cluster","p_adj_age:cluster")

colnames(ANOVA_post_hoc) = colnames_2save

# correct for multiple comparison
size =  3 
sub_poss1 = c("high","high","med")
sub_poss2 = c("med","low","low")

for (i in (1:4)) {
  col2use = var2use[i]
  print(col2use)
  
  # post hoc done by hand!
  for(o in 1:3){
    
    sub1 = sub_poss1[o]
    substypeA =  subset(data2stat,data2stat$cluster==sub1)
    #for (d in 1:3){
    
    sub2 = sub_poss2[o]
    substypeB=  subset(data2stat,data2stat$cluster==sub2)
    # define vars
    
    title= title_list[i]
    x_var = 'mullen_ageMo'
    y_var = col2use
    subgrp_var = "VABSnorms_cluster"
    maxIter = 500
    df2use = rbind(substypeA,substypeB)
    
    # define the formula
    form2use = as.formula(sprintf("%s ~ %s*%s + (%s|subjectId)", #(1|%s) + 
                                  y_var, x_var, subgrp_var, x_var))# "dataset",
    
    ctrl = lmerControl(optCtrl=list(maxfun=maxIter), # lmerControl(optCtrl=list(maxfun=maxIter)
                       check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4),check.nobs.vs.nRE = "ignore")
    # run the model
    m_allgrps <- eval(substitute(lmer(formula = form2use, data = df2use, na.action = na.omit, control = ctrl)))
    
    
    # saving 
    lme2save = anova(m_allgrps)
    lme2save$`F value` = round(lme2save$`F value`, digits = 2)
    lme2save$`DenDF` = round(lme2save$`DenDF`, digits = 2)
    lme2save$`Sum Sq`= round(lme2save$`Sum Sq`, digits = 2)
    lme2save$`Mean Sq` = round(lme2save$`Mean Sq`, digits = 2)
    
    for (j in 1:3){
      if (lme2save[j,"Pr(>F)"]<= 0.001){
        lme2save[j,"p-value_new"] = '<0.001'
      }else if (lme2save[j,"Pr(>F)"]> 0.001){
        lme2save[j,"p-value_new"] = paste("=",as.character(round(lme2save[j,"Pr(>F)"],digits = 3),sep = ""))}
      
    }
    
    #correct p value for multiple comparison
    for (a in 1:3){
      p = lme2save[a,"Pr(>F)"]
      p_adj_ = p.adjust(p, method = "fdr", n = size)
      lme2save[a,'p_adj_temp'] = p_adj_
    }
    
    #change p-adj format to be readable in excel
    for (d in 1:3){
      if (lme2save[d,"p_adj_temp"]<= 0.001){
        lme2save[d,"p_adj"] = '<0.001'
      }else if (lme2save[d,"p_adj_temp"]> 0.001){
        lme2save[d,"p_adj"] = paste("=",as.character(round(lme2save[d,"p_adj_temp"],digits = 3),sep = ""))}
      
    }
    #print(lme2save)
    #save the post-hoc lme complete
    lm = lme2save[,c("Sum Sq","Mean Sq","NumDF","DenDF","F value","p-value_new","p_adj")]
    colnames(lm) = c("Sum Sq","Mean Sq","NumDF","DenDF","F value","p-value","p_adj")
    
    write.csv(lm,file.path(result_path,"post_hoc",'MSEL_UCSD_VABSnorm',paste('Post_hoc_ANOVA_lme_spagh2_UCSD_VABSnorm_',col2use,"_",sub1,"_",sub2,".csv",sep='')))
    
    #save in a unique table
    vect2save = c(col2use,sub1,sub2,as.list(as.vector(t(lm))))
    ANOVA_post_hoc = rbind(ANOVA_post_hoc,vect2save) 
  }
}



write.csv(ANOVA_post_hoc[2:length(ANOVA_post_hoc$col2use),],file.path(result_path,'Posthoc_USCD_VABSnorm_MSELlong.csv'))

```


### model comparison in for loop for saving AIC matrix
```{r,message=FALSE,warning=FALSE, errors= FALSE}
# define which data will be used
# data2stat = data4normativemodel[data4normativemodel$VABSnorms_cluster=="high" | data4normativemodel$VABSnorms_cluster=="med" |data4normativemodel$VABSnorms_cluster=="low",]
# data2stat$VABSnorms_cluster <- factor( data2stat$VABSnorms_cluster,levels = c("high","med","low"));
sub_list = c("high","med","low")

# prepare model matrix with name of the model and the name of the columns in the DB
models2use = c("VABSnorm","REVAL")
colnames= c('VABSnorms_cluster','cluster')
model_list = c('VABSnorms_cluster','cluster')

# prepare AIC matrix
AIC_matrix = as.data.frame (matrix(nrow=length(models2use),ncol= 4))
colnames(AIC_matrix)= var2use
rownames(AIC_matrix) = models2use


# loop trought the Mullen Varibales
for (i in 1:length(var2use)){
  col2use = var2use[i]
  # loop trought the models (REVAL, VABS_norms)
  for (j in 1: length(models2use)){
    sub2use = model_list[j]
    resp = spaghettiPlot2_nodataset(
      data2use = data2stat,
      x_var = 'mullen_ageMo',
      y_var = col2use,
      subgrp_var = sub2use,
      cov_vars=NULL,
      xLabel = 'age',
      yLabel = "age equivalent",
      modelType = "linear",
      fname2save = NULL,
      plot_dots = FALSE, plot_lines = TRUE,
      dot_alpha = 1/10, line_alpha = 3/10,
      xLimits = c(0, 60), yLimits = c(0, 90),
      lineColor = "cluster", dotColor = "cluster",
      legend_name = "Slope", maxIter = 500,
      plot_title = title
    )
    
    AIC_matrix[j,i]= AIC(resp$lme_model)
    print(paste(sub2use,col2use ))
    print(AIC(resp$lme_model))
  }
}

# save AIC_matrix
# write.csv(AIC_matrix,file.path(result_path,'AIC_matrix_REVAL_VABSnorm.csv'))
```

### saving cluster labels to be plotted for umap
uncomment here if you want to save colusters labels for having some nice visualization on UMAP
```{r, include=FALSE}
#save for UMAP
#col2save = c('subjectId','gender', 'vine_ComTotal_DomStd','vine_DlyTotal_DomStd','vine_SocTotal_DomStd',
#             'vine_MtrTotal_DomStd','time',"cluster" ,"VABSnorms_cluster"  )
#data2save= data4normativemodel[data4normativemodel$time==1,col2save]
#write.csv(data2save,file.path(result_path,'clusters4UMAP.csv'))

# save for other things
#write.csv(data4normativemodel,file.path(result_path,'uscd_4modelcluster_longitudinal.csv'))
```