---
title: "R Notebook"
author: "Veronica Mandelli"
output:
  html_document:
    df_print: paged
# VISUALIZE REVAL RESULTS ON OLDER COHORT
this script analyzes the results from reval clustering on NDA dataset older cohort. 
It help visualize: stratification model reproducibility and clusters'characteristics.

### load libraries, data and set paths
```{r, message=FALSE, warning=FALSE,echo=TRUE}
#load libraries
library(gridExtra)
library(ggplot2)
library(reshape2)
library(sigclust)
library(dplyr)
library(ggpubr)

#set working directories
work_dir = '/Users/vmandelli/OneDrive - Fondazione Istituto Italiano Tecnologia/vine_old'
plot_path = file.path(work_dir,'plot')
data_path = file.path(work_dir,'results')
tidy_data_path = file.path(work_dir,'data','tidy','vineland_21022')
results_path = file.path(work_dir,'results')
code_path = file.path(work_dir,'code')
source(file.path(code_path,"cohens_d.R")) 

# data
datafile = 'VABS_old_withcl_280722_withgridsearch.csv'
data <- read.csv(file.path(results_path,datafile))
data$age_years =  round(data$interview_age/12)
```


### change cluster labels in a more easy-to-read way
```{r,message=FALSE, warning=FALSE}

# for 3 cluster solution
  for (i in (1:length(data$cluster_domain))) {
    if (data[i, 'cluster_domain'] == '0'){
      data[i,'clust_dom'] = "high"
    }
    else if (data[i, 'cluster_domain'] == '1'){
      data[i,'clust_dom'] = "low"
    }
    else if (data[i, 'cluster_domain'] == '2'){
      data[i,'clust_dom'] = "very_low"
      }
  }
#------------------------------------------------------------------------------------------------------------#
# drop the low group
#data=data[data$clust_dom_3sol=="high"|data$clust_dom_3sol=="low",]
```


### run this cell if you want to run the subsequent analysis only on the high and low subtype
```{r,message=FALSE, warning=FALSE}
# exlcude very_low subtype
#data=data[data$clust_dom %in% c('high','low'),]
```

### print out some numbers
```{r,message=FALSE, warning=FALSE,echo=TRUE}
print(paste('total sample size',length(data$subjectkey)))
print(paste('total number in high group',length(data[data$clust_dom == 'high','subjectkey'] )))
print(paste('total number in high group',length(data[data$clust_dom == 'low','subjectkey'] )))
print(paste('total number in high group',length(data[data$clust_dom == 'very_low','subjectkey'] )))

print(table(data$clust_dom,data$TR_TS))
```

### produce scatter plot for the input features - VABS domain standard scores
```{r,message=FALSE, warning=FALSE}
# define colors
# pink = c('#980043',"#DD1C77","#C994C7","#E7E1Ef")
# blue = c('#08519c',"#3182bd","#9ecae1","#deebf7")
# green = c('#006d2c','#31a354','#a1d99b','#e5f5e0')
# yellow = c('#ff9900','#ffcc33','#ffff00','#ffff99')
pink = c('#980043',"#C994C7","#E7E1Ef")
blue = c('#08519c',"#9ecae1","#deebf7")
green = c('#006d2c','#a1d99b','#e5f5e0')
yellow = c('#ff9900','#ffff00','#ffff99')
col_list = list(pink,blue,green,yellow)

#color = c("#FF9933","#3399CC","#009900","#CC0099") # old color

# set some useful lists of variables
dom_list = c('high','med','low')
var_list = c('communicationdomain_totalb', 'livingskillsdomain_totalb', 'socializationdomain_totalb')
title_list = c('communication', 'livingskills', 'socialization')

# initialize some other useful variables
title = title 
data2use = data[data$TR_TS=="TR",]
plt_tr <- list()


# looping on VABS variables to 
for (j in (1:3)){
  #define varible, color to use and titels to the plots
  var2use = var_list[j]
  col_name = col_list[j]
  color = col_list[[j]] 
  title = title_list[j]
  
  p = ggplot(data = data2use, 
             aes_string(x = 'clust_dom',
                        y = var2use,
                        color='clust_dom')) 
  p = p + geom_jitter() + geom_boxplot(fill = NA, colour = "#000000", outlier.shape = NA)
  p = p + scale_colour_manual(values = color)
  p = p + guides(colour = FALSE)
  p = p + xlab("") + ylab("")
  p = p + ggtitle(title_list[j]) 
  p = p + theme_bw() 
  p = p + theme(plot.title = element_text(hjust = 0.5))
  ## changed here
  p = p + theme(axis.text.x = element_text(face="bold", size=10))
  p = p + theme(axis.text.y = element_text(face="bold", size=10))
  p = p + scale_y_continuous( limits=c(10, 120))
  p = p + theme(axis.text=element_text(size=10),
                axis.title=element_text(size=8,face="bold"))
  
  print(p)
  # evantually save the plot
  #ggsave(paste(plot_path,'/',db_type,'_total_',var2use,".pdf", sep=""))
  plt_tr[[j]] = p
}
########################## TEST
data2use = data[data$TR_TS=="TS",]
plt_ts <- list()


# looping on VABS variables to 
for (j in (1:3)){
  #define varible, color to use and titels to the plots
  var2use = var_list[j]
  col_name = col_list[j]
  color = col_list[[j]] 
  title = title_list[j]
  
  p = ggplot(data = data2use, 
             aes_string(x = 'clust_dom',
                        y = var2use,
                        color='clust_dom')) 
  p = p + geom_jitter() + geom_boxplot(fill = NA, colour = "#000000", outlier.shape = NA)
  p = p + scale_colour_manual(values = color)
  p = p + guides(colour = FALSE)
  p = p + xlab("") + ylab("")
  p = p + ggtitle(title_list[j]) 
  p = p + theme_bw() 
  p = p + theme(plot.title = element_text(hjust = 0.5))
  ## changed here
  p = p + theme(axis.text.x = element_text(face="bold", size=10))
  p = p + theme(axis.text.y = element_text(face="bold", size=10))
  p = p + scale_y_continuous( limits=c(10, 120))
  p = p + theme(axis.text=element_text(size=10),
                axis.title=element_text(size=8,face="bold"))
  
  print(p)
  # evantually save the plot
  #ggsave(paste(plot_path,'/',db_type,'_total_',var2use,".pdf", sep=""))
  plt_ts[[j]] = p
}


# bind together

g = grid.arrange(
  plt_tr[[1]],
  plt_ts[[1]],
  plt_tr[[3]],
  #plt_tr[[4]],
  plt_ts[[3]],
  plt_tr[[2]],
  plt_ts[[3]],
 # plt_ts[[4]],
  nrow = 3)
g
#ggsave(file.path(plot_path,paste('OLD_VABS_description_withverylow_ver.png',sep="")), g, width = 5 ,height = 10)
```

### plot REVAL stability plot
```{r,message=FALSE, warning=FALSE}
reval_val = read.csv(file.path(results_path,"val_metric_for_plot_280722.csv"))
reval_val$cat = 'validation'

reval_tr = read.csv(file.path(results_path,"train_metric_for_plot_280722.csv"))
reval_tr$cat = 'train'
mat = rbind(reval_tr,reval_val)
colors = c("#cc0000","#000099")
# spaghetti plot
gg.spaghetti_x<-ggplot(mat, aes (x = k, y =value , group = cat, color = cat))
gg.spaghetti_x <-gg.spaghetti_x +geom_line() + geom_point()   +
                                 geom_errorbar(aes(ymin= value - CI, ymax= value + CI), width = 0.1, size = 0.5) + 
                                 theme(axis.text.x = element_text(size=10), axis.text.y = element_text(size=10)) +
                                 scale_color_manual(values=colors)+theme_classic()
                                  

#ggsave(file.path(plot_path,'REVAL_stability.png'),width = 6,height = 4)
gg.spaghetti_x
```


### uncomment here if you want more information about sex and age distribution
```{r, warning=FALSE,error=FALSE, include = FALSE}
#are there any differencesc in AGE at T1
clsol_2use = 'clust_dom'
color = c("#FF9933","#3399CC","#009900")
### old color "#FF6666","#99CC66","#99FFFF","#E7E1Ef")
                    
data2plot = data
#data2plot$clust_dom <- factor( data2plot$clust_dom,levels = c('high','med','low'));

data2plot$sex <- factor( data2plot$sex,levels = c("F","M"))
# age plot
p = ggplot(data =data2plot ,aes_string(x = clsol_2use, y ='age_years', color=clsol_2use) )
p = p + geom_jitter() + geom_boxplot(fill = NA, colour = "#000000", outlier.shape = NA)
p = p + scale_colour_manual(values = color)
p = p + guides(colour = FALSE)
p = p + scale_colour_manual(values = color)
p = p + xlab("") + ylab('months')
p = p + ggtitle('age')
p = p + theme_bw()
p = p + theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))
p = p + theme(axis.text.x = element_text(face="bold", size=10))
p = p + theme(axis.text.y = element_text(face="bold", size=10))
#p = p + scale_y_continuous( limits=c(72,))
#ggsave(paste(plot_path,'/NDAR',db_type,'_ageT1.pdf', sep=""))
p
# sex plot
sex =ggplot(na.omit(subset(data2plot, select = c(sex,clust_dom))), aes(x= sex,  group=clust_dom)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..),
                   y= ..prop.. ), stat= "count", vjust = -.5) +
    labs(y = "Percent",fill = "gender") +
    facet_grid(~clust_dom) +
    scale_y_continuous(labels = scales::percent)+ ylim(0,0.87)
#sex = sex + ggtitle(paste("NDAR_",db_type,"VABS-cluster",sep=''))
sex = sex + ggtitle('Gender distribution')
sex = sex  + theme_bw()
sex = sex  + theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))
sex = sex  + theme(axis.text.x = element_text(face="bold", size=10))
sex = sex  + theme(axis.text.y = element_text(face="bold", size=10))
sex
#ggsave(file.path(plot_path,paste('NDAR',db_type,'_sex.pdf', sep="")))

lay =rbind(c(1,1,2,2,2,2))

#pdf(paste(plot_path,'/age_sex',db_type,'.pdf', sep=""))
new_grid = grid.arrange(
     p,sex,
      layout_matrix = lay)

#dev.off()
new_grid
ggsave(file.path(plot_path,'gender_age_old_subtyping.png'),new_grid,width = 6,height = 4)


### stats
# gender
gender_table =  table(data$sex,data$clust_dom)
chisq <- chisq.test(t(gender_table))
print(chisq)
print(gender_table)
# pairwise comparison
sub_1_poss = c('high','high','low')
sub_2_poss = c('low','very_low','very_low')
gendet_table_sum =  as.data.frame(matrix(nrow=3,ncol=5))
colnames(gendet_table_sum) = c( 'sub1','sub2','Xsquared','df','pvalue')
for (d in 1:3){
  list_=c(sub_1_poss[d],sub_2_poss[d])
  subset_ = subset(data,data$clust_dom %in% list_)
  subset_$clust_dom = factor(subset_$clust_dom, levels = list_)
  gender_table_ = table(subset_$clust_dom,subset_$sex)
  hisq <- chisq.test(t(gender_table_))
  print(gender_table_)
  print(hisq)
  gendet_table_sum[d,'sub1']=sub_1_poss[d]
  gendet_table_sum[d,'sub2']=sub_2_poss[d]
  gendet_table_sum[d,'Xsquared']=hisq$statistic
  gendet_table_sum[d,'df']=hisq$parameter
  gendet_table_sum[d,'pvalue']=hisq$p.value
}
write.csv(gendet_table_sum,file.path(results_path,'NDA_gender_diff_tab.csv'))

# age 
#df = subset(data,data$TR_TS=="TS")
#age_table = table(df$clust_dom,df$age_years)
#write.csv(age_table,file.path(results_path,'NDA_age_distribution_agebyclusters_TS.csv'))

# pairwise
sub_1_poss = c('high','high','low')
sub_2_poss = c('low','very_low','very_low')
age_table_sum =  as.data.frame(matrix(nrow=3,ncol=7))
colnames(age_table_sum) = c( 'sub1','sub2','tstat','df','pvalue',"mean_sub1","mean_sub2")
for (d in 1:3){
  x= subset(data,data$clust_dom ==sub_1_poss[d])
  y= subset(data,data$clust_dom ==sub_2_poss[d])
  t_res <- t.test(x[,"age_years"],y[,"age_years"])
  print(t_res)
  age_table_sum[d,'sub1']=sub_1_poss[d]
  age_table_sum[d,'sub2']=sub_2_poss[d]
  age_table_sum[d,'tstat']=t_res$statistic
  age_table_sum[d,'df']=t_res$parameter
  age_table_sum[d,'pvalue']=t_res$p.value
  age_table_sum[d,'mean_sub1']=t_res$estimate[1]
  age_table_sum[d,'mean_sub2']=t_res$estimate[2]
}
write.csv(age_table_sum,file.path(results_path,'NDA_age_diff_tab_old.csv'))

summary_age_old = as.data.frame(matrix(nrow=4, ncol = 5))
colnames(summary_age_old) = c("mean","sd","median","max","min")
rownames(summary_age_old) = c('total','high','low','very_low')
for (sub in c('high','low','very_low')){
  summary_age_old[sub,'mean'] = mean(data[data$clust_dom==sub,'age_years'])
  summary_age_old[sub,'sd'] = sd(data[data$clust_dom==sub,'age_years'])
  summary_age_old[sub,'median'] = median(data[data$clust_dom==sub,'age_years'])
  summary_age_old[sub,'max'] = max(data[data$clust_dom==sub,'age_years'])
  summary_age_old[sub,'min'] = min(data[data$clust_dom==sub,'age_years'])
}

  summary_age_old['total','mean'] = mean(data$age_years)
  summary_age_old['total','sd'] = sd(data$age_years)
  summary_age_old['total','median'] = median(data$age_years)
  summary_age_old['total','max'] = max(data$age_years)
  summary_age_old['total','min'] = min(data$age_years)
  
write.csv(summary_age_old,file.path(results_path,'NDA_summary_age.csv')) 
```

### uncomment some info about cluster very low
```{r,include = FALSE,message=FALSE, warning=FALSE}
# cl_verylow = subset(data,data$clust_dom=='very_low')
# hist(cl_verylow$ageyrs_round)
# hist(cl_verylow$composite_totalb)
# table(cl_verylow$collection_id)
```

### uncomment if you want to check subject ids
```{r}
# save sub_ids with collection_id
#data_subids = data[,c('subjectkey',"collection_id")]
#write.csv(data_subids,file.path(results_path,'subid_collectionid_ndaold.csv'))
```

### check cluster significance - run sigclust
```{r}
#data
umap_datapath = file.path(work_dir,"results")
plot_path = file.path(work_dir,"plot")
filename = "UMAP_values_VABSold_280722.csv"
data <- read.csv(file.path(umap_datapath,filename), header=TRUE)
colnames(data)= c('subjectkey','UMAP_1','UMAP_2','TR_TS','cluster')


# run sigclust
# set option
nreps_sim = 10000
nreps_sigclust_repeat = 10000

data$cluster =factor(data$cluster, levels = c("2",'0','1'))
# run sigclust on train
data2use_tr = data %>% filter(TR_TS=="tr") %>% select(UMAP_1,UMAP_2,cluster)
rseo_sigclust_res_train = sigclust(data2use_tr[,c("UMAP_1","UMAP_2")],
                             nsim = nreps_sim, nrep = nreps_sigclust_repeat,
                             labflag = 1, 
                             label = data2use_tr[,"cluster"],
                             icovest=2)
p_train = ggplot(data=data2use_tr, aes(x=UMAP_1,y=UMAP_2)) +
    geom_density_2d_filled() + 
    geom_point(data=data2use_tr, aes(colour=cluster), alpha = 0.3,size = 0.5)+ggtitle('train')+ 
    scale_y_continuous( limits=c(-11, 11)) + scale_x_continuous(limits=c(-7,5))  + theme_bw() 
p_train
#rseo_sigclust_res_train
p_val_train = (sum(rseo_sigclust_res_train@simcindex<=rseo_sigclust_res_train@xcindex)+1)/(nreps_sim+1)
p_val_train

# run sigclust on test
data2use_ts = data %>% filter(TR_TS=="ts") %>% select(UMAP_1,UMAP_2,cluster)
rseo_sigclust_res_test = sigclust(data2use_ts[,c("UMAP_1","UMAP_2")],
                             nsim = nreps_sim, nrep = nreps_sigclust_repeat,
                             labflag = 1, 
                             label = data2use_ts[,"cluster"],
                             icovest=2)
#data2use_ts =  data[,c('UMAP_1','UMAP_2','cluster')]

p_test = ggplot(data=data2use_ts, aes(x=UMAP_1,y=UMAP_2)) +
    geom_density_2d_filled() + 
    geom_point(data=data2use_ts, aes(colour=cluster), alpha = 0.3,size = 0.5)+ggtitle('validation') + 
    scale_y_continuous( limits=c(-11, 11)) + scale_x_continuous(limits=c(-7,5))  + theme_bw() 
p_test


#rseo_sigclust_res_test
p_val_test = (sum(rseo_sigclust_res_test@simcindex<=rseo_sigclust_res_test@xcindex)+1)/(nreps_sim+1)
p_val_test


# save outputs
umap_tr_ts = ggarrange(p_train, p_test, nrow=1, common.legend = TRUE, legend="bottom")
umap_tr_ts
#ggsave(file.path(plot_path,'UMAP_train_290722.png'), p_train, width = 7 ,height = 6)
#ggsave(file.path(plot_path,'UMAP_test_290722.png'), p_test, width = 7 ,height = 6)
```



